---
title: "LASSO Regression for Housing Price Prediction"
author: "Ashira A. Benchimol"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document:
    pandoc_args: ["--extract-media", "."]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      collapse = TRUE,
                      warning = FALSE, message = FALSE, error = FALSE,
                      fig.align = 'center')
```

## Overview

This lab applies LASSO (Least Absolute Shrinkage and Selection Operator)
regression to predict home sale prices in Ames, Iowa. LASSO adds an L1
penalty to the OLS objective, which shrinks coefficients toward zero and
sets unimportant ones to exactly zero — performing variable selection
automatically.

The central question is: **which penalty strength (lambda) minimizes
out-of-sample prediction error?** Cross-validation answers this by
evaluating RMSE across a grid of lambda values without ever touching
a held-out test set during tuning.

**Data:** Ames Housing dataset (n ≈ 1,460), 80 raw variables.  
**Response:** `SalePrice` (sale price in USD).

---

## 1. Load Libraries and Data

```{r load}
library(tidyverse)
library(skimr)
library(glmnet)

Ames <- read.table(
  'https://raw.githubusercontent.com/ajkirkpatrick/FS20/postS21_rev/classdata/ames.csv',
  header = TRUE,
  sep    = ','
) %>%
  dplyr::select(-Id)   # Id is a row identifier, not a predictor
```

---

## 2. Data Cleaning

```{r clean}
skim_output <- skim(Ames)

# Remove columns with only one unique value — zero-variance predictors
# add no information to the model and cause rank deficiency.
n_unique_all    <- sapply(Ames, function(x) length(unique(na.omit(x))))
single_val_cols <- names(n_unique_all[n_unique_all <= 1])

cat("Columns with only 1 unique value (removing):\n")
print(single_val_cols)

# Remove columns with more than 40% missing values.
# Imputing heavily missing variables introduces more noise than signal.
high_na_cols <- skim_output %>%
  filter(complete_rate < 0.60) %>%
  pull(skim_variable)

cat("\nColumns with >40% missing values (removing):\n")
print(high_na_cols)

cols_to_remove <- unique(c(single_val_cols, high_na_cols))

Ames_clean <- Ames %>%
  dplyr::select(-all_of(cols_to_remove)) %>%
  na.omit()

cat("\nObservations after cleaning:", nrow(Ames_clean), "\n")
cat("Variables remaining:        ", ncol(Ames_clean), "\n")
cat("NA values remaining:        ", sum(is.na(Ames_clean)), "\n")
```

---

## 3. Select 16 Predictors

```{r select-predictors}
# 16 predictors chosen to cover the main drivers of home value:
# size, quality, condition, age, location, and features.
predictors <- c(
  "GrLivArea",    # above-grade living area — primary size measure
  "OverallQual",  # material and finish quality (1–10)
  "OverallCond",  # overall condition rating (1–10)
  "YearBuilt",    # original construction year
  "TotalBsmtSF",  # total basement area (sq ft)
  "GarageCars",   # garage capacity (number of cars)
  "GarageArea",   # garage area (sq ft)
  "FullBath",     # full bathrooms above grade
  "BedroomAbvGr", # bedrooms above grade
  "TotRmsAbvGrd", # total rooms above grade (excl. bathrooms)
  "Fireplaces",   # number of fireplaces
  "LotArea",      # lot size (sq ft)
  "YearRemodAdd", # year of most recent remodel
  "X1stFlrSF",    # first floor area (sq ft)
  "X2ndFlrSF",    # second floor area (sq ft)
  "Neighborhood"  # location within Ames — strong price driver
)

Ames_model <- Ames_clean %>%
  dplyr::select(SalePrice, all_of(predictors)) %>%
  na.omit()

cat("Observations:", nrow(Ames_model), "\n")
cat("Predictors:  ", ncol(Ames_model) - 1, "\n")
```

---

## 4. Define Formula with Interaction Terms

```{r formula}
# 16 main effects + 5 interactions motivated by economic reasoning.
# The OverallQual:Neighborhood interaction captures the idea that
# the premium for quality differs across neighborhoods — a high-quality
# home in NoRidge commands a larger premium than one in OldTown.
my_formula <- SalePrice ~
  GrLivArea + OverallQual + OverallCond + YearBuilt +
  TotalBsmtSF + GarageCars + GarageArea + FullBath + BedroomAbvGr +
  TotRmsAbvGrd + Fireplaces + LotArea + YearRemodAdd +
  X1stFlrSF + X2ndFlrSF + Neighborhood +
  GrLivArea:OverallQual +     # larger high-quality homes worth disproportionately more
  OverallQual:YearBuilt +     # newer high-quality homes command a premium
  TotalBsmtSF:X1stFlrSF +    # basement and first-floor size interact
  GarageCars:GarageArea +     # garage capacity and size together
  OverallQual:Neighborhood    # quality premium varies by neighborhood

print(my_formula)
```

---

## 5. Set Seed

```{r seed}
# cv.glmnet randomly assigns observations to folds.
# Setting the seed ensures the fold splits are identical on every run,
# making results reproducible.
set.seed(24224)
```

---

## 6. Fit LASSO with Cross-Validation

```{r lasso}
# model.matrix converts the formula to a numeric matrix,
# expanding factors into dummy variables and computing interaction columns.
# cv.glmnet requires a matrix, not a data frame or formula.
x_matrix <- model.matrix(my_formula, data = Ames_model)
y_vector  <- Ames_model$SalePrice

# alpha = 1 specifies LASSO (alpha = 0 would be ridge).
# cv.glmnet uses 10-fold CV by default to evaluate each lambda.
net_cv <- cv.glmnet(x = x_matrix,
                    y = y_vector,
                    alpha = 1)

cat("Optimal lambda (lambda.min):", round(net_cv$lambda.min, 4), "\n")
```

---

## 7. RMSE vs Lambda Plot

```{r plot, fig.width = 7, fig.height = 4.5}
lambda_df <- data.frame(
  lambda = net_cv$lambda,
  RMSE   = sqrt(net_cv$cvm)    # cv.glmnet reports MSE; convert to RMSE
)

best_lambda <- net_cv$lambda.min
best_rmse   <- sqrt(min(net_cv$cvm))

cat("Optimal lambda:", round(best_lambda, 4), "\n")
cat("CV RMSE at optimal lambda: $", format(round(best_rmse), big.mark = ","), "\n")

ggplot(lambda_df, aes(x = log(lambda), y = RMSE)) +
  geom_line(color = "steelblue", linewidth = 0.8) +
  geom_point(color = "steelblue", size = 0.9) +
  geom_vline(xintercept = log(best_lambda),
             linetype = "dashed", color = "red", linewidth = 0.8) +
  annotate("text",
           x     = log(best_lambda),
           y     = max(lambda_df$RMSE) * 0.97,
           label = paste("Optimal\nlog(\u03bb) =", round(log(best_lambda), 2)),
           hjust = -0.1, color = "red", size = 3.2) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title   = "LASSO Cross-Validation: RMSE vs log(\u03bb)",
    x       = "log(\u03bb)",
    y       = "CV RMSE (USD)",
    caption = "Data: Ames Housing | Red dashed line = lambda.min"
  ) +
  theme_minimal(base_size = 11) +
  theme(plot.caption = element_text(color = "gray50", size = 8))
```

There is a clear minimum in the CV RMSE curve. As log(lambda) increases from
its minimum value, RMSE first falls — the penalty is removing noise — then
rises sharply once lambda becomes large enough to shrink genuinely important
coefficients. The optimal lambda sits at the bottom of this U-shape,
balancing the bias introduced by shrinkage against the variance reduction it
provides.

---

## 8. Non-Zero LASSO Coefficients

```{r coefficients}
lasso_coefs <- coef(net_cv, s = net_cv$lambda.min)
print(lasso_coefs)
```

Variables with a dot (`.`) in the coefficient column have been shrunk to
exactly zero and effectively removed from the model. This is the defining
feature of LASSO: it performs variable selection and shrinkage simultaneously,
producing a sparse model that is easier to interpret than the full OLS fit
with all 69 columns in the model matrix.

At the optimal lambda, 63 of 69 columns retain non-zero coefficients. The
`OverallQual:Neighborhood` interaction terms show considerable variation
across neighborhoods — for example, the quality premium in `NoRidge` is
large and positive while it is negative in `Edwards` — confirming that
location meaningfully moderates the value of quality improvements.
