---
title: "Logistic Regression for Bank Subscription Prediction"
author: "Ashira A. Benchimol"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document:
    pandoc_args: ["--extract-media", "."]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      collapse = TRUE,
                      warning = FALSE, message = FALSE, error = FALSE,
                      fig.align = 'center')
```

## Overview

This lab fits a series of logistic regression models to predict whether a
bank client subscribed to a term deposit following a phone marketing
campaign. The dataset is heavily imbalanced — roughly 89% of clients did
not subscribe — which makes accuracy alone a misleading metric. Sensitivity
(the fraction of actual subscribers correctly identified) and specificity
(the fraction of non-subscribers correctly identified) are more informative
and are tracked across all models.

Four models are compared, each adding predictors or an interaction term.
The goal is to evaluate whether additional predictors improve classification
beyond what `duration` alone provides.

**Data:** Portuguese bank marketing dataset (n = 4,521).  
**Response:** `y` — whether the client subscribed (yes = 1, no = 0).

---

## 1. Load Data and Prepare Response Variable

```{r load}
library(tidyverse)
library(caret)

# Load the bank dataset; drop the separate 'default' column
bank <- read.table("https://ec242.netlify.app/data/bank.csv",
                   header = TRUE,
                   sep    = ",") %>%
  dplyr::select(-default)

# Convert the character response to a binary integer (1 = subscribed, 0 = not)
bank <- bank %>%
  dplyr::mutate(y = case_when(
    y == "yes" ~ as.integer(1),
    y == "no"  ~ as.integer(0)
  ))

cat("Subscription class distribution:\n")
print(table(bank$y))
cat(sprintf("Subscription rate: %.1f%%\n", mean(bank$y) * 100))
```

The subscription rate is approximately 11.5%, confirming heavy class
imbalance. A naive classifier that always predicts "no" would achieve
~88.5% accuracy without learning anything — so accuracy alone is not
sufficient to evaluate these models.

---

## 2. Check for Missing Values

```{r missing}
cat("Total NA values:", sum(is.na(bank)), "\n")
bank <- na.omit(bank)
cat("Observations after NA removal:", nrow(bank), "\n")
```

---

## 3. Train / Test Split (80/20)

```{r split}
set.seed(24224)

num_obs     <- nrow(bank)
train_index <- sample(num_obs, size = trunc(0.80 * num_obs))

train_data <- bank[train_index, ]
test_data  <- bank[-train_index, ]

cat("Training observations:", nrow(train_data), "\n")
cat("Test observations:    ", nrow(test_data),  "\n")
```

---

## 4. Fit Four Logistic Regression Models

Models are built cumulatively, adding predictors in order of expected
predictive value based on domain knowledge about phone marketing campaigns.

```{r models}
# Model 1: duration only
# Call duration is the strongest single predictor — a longer call
# suggests a client engaged seriously with the offer.
model_1 <- glm(y ~ duration,
               data = train_data, family = "binomial")

# Model 2: add account balance
# Clients with higher balances may be more receptive to savings products.
model_2 <- glm(y ~ duration + balance,
               data = train_data, family = "binomial")

# Model 3: add age
# Age may proxy for life stage and financial priorities.
model_3 <- glm(y ~ duration + balance + age,
               data = train_data, family = "binomial")

# Model 4: add campaign contacts and a duration:balance interaction
# More contacts in a campaign may indicate follow-up on interested clients.
# The interaction allows the effect of call duration to vary with balance.
model_4 <- glm(y ~ duration + balance + age + campaign + duration:balance,
               data = train_data, family = "binomial")
```

---

## 5. Evaluate on Training and Test Data

```{r evaluate}
# Predict class labels from estimated probabilities using a 0.5 threshold.
get_logistic_pred <- function(mod, data, pos = 1, neg = 0, cut = 0.5) {
  probs <- predict(mod, newdata = data, type = "response")
  ifelse(probs > cut, pos, neg)
}

models      <- list(model_1, model_2, model_3, model_4)
model_names <- paste0("Model ", 1:4)

# Compute confusion matrices for all four models on both datasets
train_cms <- lapply(models, function(m) {
  preds <- get_logistic_pred(m, data = train_data)
  tab   <- table(predicted = preds, actual = train_data$y)
  confusionMatrix(tab, positive = "1")
})

test_cms <- lapply(models, function(m) {
  preds <- get_logistic_pred(m, data = test_data)
  tab   <- table(predicted = preds, actual = test_data$y)
  confusionMatrix(tab, positive = "1")
})

# Print each test confusion matrix
for (i in seq_along(models)) {
  cat(sprintf("\n--- %s | TEST ---\n", model_names[i]))
  print(test_cms[[i]])
}
```

---

## 6. Summary Table

```{r summary-table}
results <- data.frame(
  Model             = model_names,
  Train_Accuracy    = sapply(train_cms, function(cm) cm$overall["Accuracy"]),
  Test_Accuracy     = sapply(test_cms,  function(cm) cm$overall["Accuracy"]),
  Train_Sensitivity = sapply(train_cms, function(cm) cm$byClass["Sensitivity"]),
  Test_Sensitivity  = sapply(test_cms,  function(cm) cm$byClass["Sensitivity"]),
  Train_Specificity = sapply(train_cms, function(cm) cm$byClass["Specificity"]),
  Test_Specificity  = sapply(test_cms,  function(cm) cm$byClass["Specificity"])
)

print(round(results, 4), row.names = FALSE)
```

---

## 7. Interpretation

Accuracy is stable at approximately 88–89% across all four models. This is
largely explained by the class imbalance: a model that predicts "no
subscription" for every observation would achieve ~88.5% accuracy, so high
accuracy here does not indicate strong performance.

Sensitivity — the fraction of actual subscribers correctly identified — is
the more important metric for a bank trying to target potential customers.
It sits around 14–15% across models, meaning the models miss the large
majority of subscribers. This is a natural consequence of the 0.5 threshold
combined with heavy imbalance: the model assigns most observations a
predicted probability below 0.5, defaulting to "no."

Specificity is very high (~98–99%), reflecting that the models are
conservative — they rarely predict a subscription that does not occur.

Adding predictors beyond `duration` provides only marginal improvement,
suggesting `duration` captures most of the available signal in this dataset.
Train and test metrics are nearly identical across all models, indicating
no meaningful overfitting.

A practical next step would be to lower the classification threshold from
0.5 to something like 0.2 or 0.3. This would increase sensitivity at the
cost of specificity — a worthwhile tradeoff in a marketing context where
the cost of missing a potential subscriber is high relative to the cost of
an unnecessary call.
